Reconstructing Propositional Proofs in Type Theory

### 2017-10-25 paper/sections/metis-language-and-proofs.tex
minor fixed.

Where the \name{language} can be \name{fof} or \name{cnf}. The \name{name}
serves to identify the formula within the problem. Each formula
assumes a \name{role}, this could be an \name{axiom},
\name{conjecture}, \name{definition}, \name{plain} or an \name{hypothesis}.
The formulas include the constants \verb!$true!, negation unary operator (\verb!~!), and the binary connectives (\verb!&!, \verb!|!, \verb!=>!, \verb!<=>!),

\begin{myexample}
For instance, let us express the problem
$p\, \vdash \neg (p \wedge \neg p) ∨ (q \wedge \neg q)$
in \TPTP syntax. We begin by declaring the $p$ axiom using the \verb!axiom!
keyword. Later, we include the expected conclusion using the
\verb!conjecture! keyword.

@@

Let us consider the following \TSTP derivation using
the customized \TSTP syntax. Note that we remove
empty fields in the inference information, we shorten names as long as possible, and we delete the \verb!role! field when the role is
\verb!plain!.

\begin{verbatim}
fof(premise, axiom, p).
fof(goal, conjecture, p).
fof(sg0, p, inf(strip, [goal])).
fof(neg0, ~ p, inf(negate, [sg0])).
fof(norm0, ∼ p, inf(canonicalize, [neg0])).
fof(norm1, p, inf(canonicalize, [premise])).
fof(norm2, $false, inf(simplify, [norm0, norm1]))
cnf(refute0, $false, inf(canonicalize, [norm2])).
\end{verbatim}

@@

\caption{\Metis inference rules.}
  \begin{center}
  {\renewcommand{\arraystretch}{1.6}%
    \label{tab:agda-metis-table}
    \begin{tabular}{|@{\hspace{2mm}}l@{\hspace{2mm}}l@{\hspace{2mm}}l@{\hspace{2mm}}|}
    \hline
    \textbf{\Metis rule} & \textbf{Purpose} &\textbf{Theorem}\\ \hline
    \texttt{canonicalize}
      &Normalization of the formula
      &\ref{thm:canonicalize}
    \\
    \texttt{clausify}
      &Performs clausification
      &\ref{thm:clausify}
    \\
    \hyperlink{atp-conjunct}{\texttt{conjunct}}
      &Takes a formula from a conjunction
      &\ref{thm:conjunct}
    \\
    % \texttt{negate}
    %   &Negate the formula
    %   &\ref{thm:negate}
    % \\
    \texttt{resolve}
      &A general form of the resolution theorem
      &\ref{thm:resolve}
    \\
    \texttt{simplify}
      &Simplify definitions and theorems
      &\ref{thm:simplify}
    \\
    \texttt{strip}
      &Strip a goal into subgoals
      &\ref{thm:strip}
    \\[1ex]
    \hline
    \end{tabular}}
  \end{center}
\end{table}

>>>>

Where the \name{language} can be \name{fof} or \name{cnf}. The \name{name}
serves to identify the formula within the problem. Each formula
assumes a \name{role}, this could be an \name{axiom},
\name{conjecture}, \name{definition}, \name{plain} or an \name{hypothesis}.
The formulas include the constants \verb!$true! and \verb!$false!, negation unary operator (\verb!~!), and the binary connectives (\verb!&!, \verb!|!, \verb!=>!, \verb!<=>!) to represent $⊤, ⊥, ¬, ∧, ∨, ⇒, ⇔$ respectively.

\begin{myexample}
For instance, let us express the problem
$p\, ⊢ ¬  (p ∧ ¬ p) ∨ (q ∧ ¬ q)$
in \TPTP syntax. We begin by declaring the $p$ axiom using the \verb!axiom!
keyword. Later, we include the expected conclusion using the
\verb!conjecture! keyword.

@@

Let us consider the following \TSTP derivation using
the customized \TSTP syntax. Note that we remove
empty fields in the inference information, we shorten names as long as possible, we delete the \verb!role! field when the role is
\verb!plain! and we use the math symbols to express the formulas.

\begin{verbatim}
fof(premise, axiom, p).
fof(goal, conjecture, p).
fof(sg0, p, inf(strip, [goal])).
fof(neg0, ¬ p, inf(negate, [sg0])).
fof(norm0, ¬ p, inf(canonicalize, [neg0])).
fof(norm1, p, inf(canonicalize, [premise])).
fof(norm2, ⊥, inf(simplify, [norm0, norm1]))
cnf(refute0, ⊥, inf(canonicalize, [norm2])).
\end{verbatim}

@@

\caption{\Metis inference rules.}
  \begin{center}
  {\renewcommand{\arraystretch}{1.6}%
    \label{tab:agda-metis-table}
    \begin{tabular}{|@{\hspace{2mm}}c@{\hspace{2mm}}l@{\hspace{3mm}}l@{\hspace{3mm}}|}
    \hline
    \textbf{Theorem} & \textbf{Purpose} &\textbf{\Metis rule}\\ \hline

       \ref{thm:strip}
      &Strip a goal into subgoals
      &\texttt{strip}    \\[1ex]

       \ref{thm:conjunct}
      &Takes a formula from a conjunction
      &\texttt{conjunct}    \\

       \ref{thm:resolve}
      &A general form of the resolution theorem
      &\texttt{resolve}    \\

       \ref{thm:clausify}
      &Performs clausification
      &\texttt{clausify}    \\

       \ref{thm:canonicalize}
      &Normalization of the formula
      &\texttt{canonicalize}    \\

       \ref{thm:simplify}
      &Simplify definitions and theorems
      &\texttt{simplify}    \\

    \hline
    \end{tabular}}
  \end{center}
\end{table}

### 2017-10-25 paper/sections/proof-reconstruction/resolve.tex
[ paper ] fixed minor typos.
[ paper ] fixing last suggestions.
minor fixed.

Logic equivalence between propositions is a major issue to justify
prover reasoning steps. Since we left out semantics to treat only the
syntax aspects of the propositions, our approach shows equivalence
by converting propositions to their conjunctive normal form,
and reordering those and the inner disjunctions to match them.
Below, we provide some lemmas to perform such reordering tasks,
we omit the proofs but they can found in~\cite{AgdaMetis}.

First, we define \fassoc$_{⬓}$ function in~(\ref{eq:fassoc})
to convert a disjunction or a conjunction into its right-associative form.
The square symbol can be the conjunction symbol or the disjunction symbol.
We use $\fassoc_{∧}$ to convert conjunctions and $\fassoc_{∨}$ for
disjunctions.

@@

\begin{myexample}\hspace{10cm}
\begin{enumerate}
\item $\fassoc_{∨}~(((p ∨ ((q ∨ r) ∨ s)) ∨ t) = p ∨ (q ∨ (r ∨ (s ∨ t)))$.
\item $\fassoc_{∧}~((((p ∧ q) ∧ r) ∧ s) ∧ t) = p ∧ (q ∧ (r ∧ (s ∧ t)))$.
\end{enumerate}
\end{myexample}

@@

\begin{mainlemma}
  \label{lem:reorder-and-or}
  If $Γ ⊢ φ$ and $ψ : \Prop$ then $Γ ⊢ \freorder_{∧∨}~φ~ψ$, where,
   \begin{align*}
      \begin{split}
      \label{eq:reorder-and-or}
      &\freorder_{∧∨} : \Prop \to \Prop \to \Prop\\
      &\freorder_{∧∨}~φ~ψ=
        \begin{cases}
          ψ, &\text{ if } φ ≡ ψ\\
          ψ, &\text{ if } ψ ≡ ψ₁ ∧ ψ₂,\ ψ₁ ≡ \freorder_{∧∨}~φ~ψ₁\\
             &\text{ and } ψ₂ ≡ \freorder_{∧∨}~φ~ψ₂\\
          φ, &\text{ if } ψ ≡ ψ₁ ∧ ψ₂\\
          \fconjunct_{∨}~φ~ψ &\text{ otherwise.}
        \end{cases}
      \end{split}
  \end{align*}
\end{mainlemma}

@@

\begin{myexample}\hspace{10cm}
\begin{verbatim}
cnf(ref4, (~ r | p | q), inf(canonicalize, [norm5])).
cnf(ref5, (p | q | r), inf(canonicalize, [norm6])).
cnf(ref6, (p | q), inf(resolve, [$cnf(r)], [ref5, ref4])).
\end{verbatim}
In the excerpt above, we apply resolution for the first two formulas,
$¬ r ∨ p ∨ q$ and $p ∨ q ∨ r$. The last line tells us the literal mentioned
for resolution is $r$. Syntactically speaking,
we can not derive neither the conclusion $p ∨ q$ in \verb!ref6! nor apply the resolution theorem with \verb!ref4! and \verb!ref5! since the formulas does not fit the pattern required.

If the scenario would have other like replacing \verb!ref5! by
\begin{verbatim}
cnf(ref5, (r | p | q), inf(canonicalize, [norm6])).
\end{verbatim}
We have could derive $(p ∨ q) ∨ (p ∨ q)$ but again, that is not the expected result.
\end{myexample}

@@

\begin{myexample}
Continuing with the problem presented in the example above,
we can use Theorem \ref{thm:resolve} to derive $Γ ⊢ p ∨ q$.
Let be $φ₁ ≡ p ∨ q ∨ r$ and $φ₂ ≡ ¬ r ∨ p ∨ q$.

>>>>

Logic equivalence between propositions is a major issue to justify
prover reasoning steps. Since we left out semantics to treat only the
syntax aspects of propositional logic, our approach shows logic equivalence
by converting propositions to their conjunctive normal form,
and reordering those and the inner disjunctions to match them.
Below, we provide Lemma~\ref{lem:reorder-or}, Lemma~\ref{lem:reorder-and}, Lemma~\ref{lem:reorder-and-or} to perform such reordering tasks,
the omitted proofs can be found in~\cite{AgdaMetis}.

First, we define \fassoc$_{⬓}$ function in~(\ref{eq:fassoc})
to convert a disjunction or a conjunction into its right-associative form.
The square symbol (⬓) can be the conjunction symbol or the disjunction symbol.
We use $\fassoc_{∧}$ to convert conjunctions and $\fassoc_{∨}$ for
disjunctions.

@@

@@

\begin{mainlemma}
  \label{lem:reorder-and-or}
  If $Γ ⊢ φ$ and $ψ : \Prop$ then $Γ ⊢ \freorder_{∧∨}~φ~ψ$, where,
   \begin{align}
      \begin{split}
      \label{eq:reorder-and-or}
      &\freorder_{∧∨} : \Prop \to \Prop \to \Prop\\
      &\freorder_{∧∨}~φ~ψ=
        \begin{cases}
          ψ, &\text{ if } φ ≡ ψ\\
          ψ, &\text{ if } ψ ≡ ψ₁ ∧ ψ₂,\ ψ₁ ≡ \freorder_{∧∨}~φ~ψ₁\\
             &\text{ and } ψ₂ ≡ \freorder_{∧∨}~φ~ψ₂\\
          φ, &\text{ if } ψ ≡ ψ₁ ∧ ψ₂\\
          \fconjunct_{∨}~φ~ψ &\text{ otherwise.}
        \end{cases}
      \end{split}
  \end{align}
\end{mainlemma}

@@

\begin{myexamplenum}\hspace{10cm}
\label{ex:resolve-tstp}
\begin{verbatim}
cnf(ref4, (~ r | p | q), inf(canonicalize, [norm5])).
cnf(ref5, (p | q | r), inf(canonicalize, [norm6])).
cnf(ref6, (p | q), inf(resolve, [$cnf(r)], [ref5, ref4])).
\end{verbatim}
In the excerpt above, we apply resolution for the first two formulas,
$¬ r ∨ p ∨ q$ and $p ∨ q ∨ r$. The last line tells us the literal mentioned
for resolution is $r$. Syntactically speaking,
we can not derive neither the conclusion $p ∨ q$ in \verb!ref6! nor apply the resolution theorem with \verb!ref4! and \verb!ref5! since the formulas does not fit the pattern required.

If the scenario would have other like replacing \verb!ref5! by
\begin{verbatim}
cnf(ref5, (r | p | q), inf(canonicalize, [norm6])).
\end{verbatim}
We have could derive $(p ∨ q) ∨ (p ∨ q)$ but again, that is not the expected result.
\end{myexamplenum}

@@

\begin{myexample}
Continuing with the problem presented in the Example~\ref{ex:resolve-tstp},
we can use Theorem \ref{thm:resolve} to derive $Γ ⊢ p ∨ q$.
Let be $φ₁ ≡ p ∨ q ∨ r$ and $φ₂ ≡ ¬ r ∨ p ∨ q$.

### 2017-10-25 paper/sections/proof-reconstruction/strip.tex
minor fixed.

\begin{verbatim}
fof(goal, conjecture, (p & r) & q)).
fof(sg0, p, inf(strip, [goal])).
fof(sg1, p => r, inf(strip, [goal])).
fof(sg2, (p & r) => q, inf(strip, [goal])).
\end{verbatim}

@@

\begin{proof}
The proof is by induction on the cases defined by the outcome of the
$\fuh_{1}$ function.
\begin{itemize}
  \item If $n = 0$, by definition in~(\ref{eq:uh-structured})
        we conclude $Γ ⊢ φ$.
  \item If $n = 1$, we apply the following theorem proved in~\cite{AgdaProp},
    \begin{equation*}
    \texttt{∧⇒\rm{-to-}⇒⇒}\ :\  Γ ⊢ (φ₁ ∧ φ₂) ⇒ φ₃ → Γ ⊢ φ₁ ⇒ (φ₂ ⇒ φ₃).
    \end{equation*}
  \item For $n > 1$, we use induction on the structure of the second
        argument.
\vskip 2mm
\begin{itemize}

\item Case $φ ≡ φ₁ ⇒ (φ₂ ⇒ φ₃)$.
\begin{equation*}
  \begin{bprooftree}
  \AxiomC{$Γ ⊢ \fuh_1~(φ₁ ⇒ (φ₂ ⇒ φ₃))~(\suc~n)$}
  \RightLabel{by~(\ref{eq:uh-structured})}
  \UnaryInfC{$Γ ⊢ \fuh_1~((φ₁ ∧ φ₂) ⇒ φ₃)~n$}
  \RightLabel{by~ind.~hyp.}
  \UnaryInfC{Γ ⊢ (φ₁ ∧ φ₂) ⇒ φ₃}
  \RightLabel{∧⇒\rm{-to-}⇒⇒}
  \UnaryInfC{Γ ⊢ φ₁ ⇒ (φ₂ ⇒ φ₃)}
  \end{bprooftree}
\end{equation*}

@@

The function that yields the conjunction of subgoals that implies the goal of
the problem in the \Metis \TSTP derivations is not a structurally recursive
function.  We present its structural recursive function
in~(\ref{eq:strip-fixed}) after applying
the bounded technique described in Section~\ref{ssec:structural-recursion}.
We based the definition of this function based on the \Metis source code.

@@

In a similar way as we define $\fuh_{cm}$ in (\ref{eq:uh-complexity}) we define
$\fstrip_{cm}$ as the complexity measure for $\fstrip₀$ function. As excepted,
the \fstrip function is defined as follows.

@@

\begin{proof}
The proof is by induction on the structure of the
formula~$φ$ by following the cases in~(\ref{eq:strip-fixed}).
We present a straightforward case with double negation, a case with
conjunction connective, and last, the case with a negated disjunction.
We refer the reader to~\cite{AgdaMetis} for the complete proof in \Agda.

@@

$\bullet$~Case $φ ≡ φ₁ ∧ φ₂$. We get a proof for each conjunct and using the
introduction rule for conjunction connective, the expected result follows.

@@

Since \Metis proves a conjecture by refutation,
to prove each subgoal, \Metis assumes the negation of it
using the \negate rule. Then, the \negate rule is
always derived from a \strip inference application.

\begin{myexample}
In the following \TSTP derivation, the three subgoal are
introduced by the \negate rule.

\begin{verbatim}
fof(goal, conjecture, (p & r & q)).
fof(sg0, p, inf(strip,[goal])).
fof(sg1, p => r, inf(strip,[goal])).
fof(sg2, (p & r) => q, inf(strip,[goal])).
fof(neg0, ~ p, inf(negate,[sg0])).
...
fof(neg1, ~ (p => r), inf(negate,[sg1])).
...
fof(neg2, ~ ((p & r) => q), inf(negate,[sg2])).
\end{verbatim}
\end{myexample}

>>>>

\begin{verbatim}
fof(goal, conjecture, (p ∧ r) ∧ q)).
fof(sg0, p, inf(strip, [goal])).
fof(sg1, p ⇒ r, inf(strip, [goal])).
fof(sg2, (p ∧ r) ⇒ q, inf(strip, [goal])).
\end{verbatim}

@@

\begin{proof}
The proof is by induction on the cases defined by the outcome of the
$\fuh_{1}$ function.
\begin{itemize}
  \item If $n = 0$, by definition in~(\ref{eq:uh-structured})
        we conclude $Γ ⊢ φ$.
  \item For $n \geq 1$, we use induction on the structure of the second
        argument.
\vskip 2mm
\begin{itemize}
\item Case $φ ≡ φ₁ ⇒ (φ₂ ⇒ φ₃)$.
\begin{equation*}
  \begin{bprooftree}
  \AxiomC{$Γ ⊢ \fuh_1~(φ₁ ⇒ (φ₂ ⇒ φ₃))~(\suc~n)$}
  \RightLabel{by~(\ref{eq:uh-structured})}
  \UnaryInfC{$Γ ⊢ \fuh_1~((φ₁ ∧ φ₂) ⇒ φ₃)~n$}
  \RightLabel{by~ind.~hyp.}
  \UnaryInfC{Γ ⊢ (φ₁ ∧ φ₂) ⇒ φ₃}
  \RightLabel{∧⇒\rm{-to-}⇒⇒}
  \UnaryInfC{Γ ⊢ φ₁ ⇒ (φ₂ ⇒ φ₃)}
  \end{bprooftree}
\end{equation*}
using the following theorem proved in~\cite{AgdaProp},
  \begin{equation*}
  \texttt{∧⇒\rm{-to-}⇒⇒}\ :\  Γ ⊢ (φ₁ ∧ φ₂) ⇒ φ₃ → Γ ⊢ φ₁ ⇒ (φ₂ ⇒ φ₃).
  \end{equation*}

@@

The \fstrip₀ function yields the conjunction of subgoals that implies the goal of the problem in the \Metis \TSTP derivations but is not a structurally recursive
function.  Instead, we present its structural recursive version
in~(\ref{eq:strip-fixed}) after applying
the bounded technique described in Section~\ref{ssec:structural-recursion}.
We based on the definition of this function from the \Metis source code.

@@

In a similar way as we define $\fuh_{cm}$ in (\ref{eq:uh-complexity}) we define
$\fstrip_{cm}$ as the complexity measure for $\fstrip₀$ function. As expected,
the \fstrip function is defined as follows.

@@

\begin{proof}
The proof is by induction on the structure of the
formula~$φ$ by following the equations in~(\ref{eq:strip-fixed}).
We present a straightforward case with double negation, a case with
conjunction connective, and last, the case with a negated disjunction.
We refer the reader to~\cite{AgdaMetis} for the complete proof in \Agda.

@@

$\bullet$~Case $φ ≡ φ₁ ∧ φ₂$. We prove $Γ ⊢ φ₁$ and $Γ ⊢ φ₂$. Using the
introduction rule for conjunction, the expected result follows.

@@

Since \Metis proves a conjecture by refutation,
to prove each subgoal, \Metis assumes the negation of it
by using the \negate rule after the \strip inference application
that introduce such a subgoal.

\begin{myexample}
In the following \TSTP derivation, note that the three subgoals
\verb!sg0!, \verb!sg1! and \verb!sg2! are assumed by the \negate
rule in \verb!neg0!, \verb!neg1! and  \verb!neg2! respectively.

\begin{verbatim}
fof(goal, conjecture, (p ∧ r ∧ q)).
fof(sg0, p, inf(strip,[goal])).
fof(sg1, p ⇒ r, inf(strip,[goal])).
fof(sg2, (p ∧ r) ⇒ q, inf(strip,[goal])).
fof(neg0, ¬ p, inf(negate,[sg0])).
...
fof(neg1, ¬ (p ⇒ r), inf(negate,[sg1])).
...
fof(neg2, ¬ ((p ∧ r) ⇒ q), inf(negate,[sg2])).
\end{verbatim}
\end{myexample}
